---
title: Stata Basics
author: Will Doyle
output: github_document
---

```{r, echo = FALSE, message = FALSE}
require(knitr)
source('../../r/stataknitrhelper.r')
opts_chunk$set(echo = FALSE, message = FALSE, comment = NA)
```

```{r, engine = 'bash'}
## run accompanying .do file to get log file for parsing
stata-se -b -q do ./stata_basics.do	 
```

```{r}
## save log file in object
lf <- 'stata_basics.log'
```

<br>

#### PURPOSE

In this class we'll continue exploring some of Stata's basic
functionality.

<br>

## A template for `do` files
All `*.do` files for this course should follow the outline provided by the
example do files. An example template is also posted on the course
website. The point is to have a standard structure that will make
sense for every new `*.do` file you create.

<br>

## Downloading packages (`*.ado` files)
One of Stata's great virtues is that it is highly extensible. Users
are frequently creating new commands for Stata and posting them on the
web. To get these commands, we use the `net search` command. From
there, a list of possible matches will be presented. After choosing
the one you want, go ahead and download it into Stata. You now have
access to the new command.


```{r}
start <- 'downloading ado files'
end <- 'load in school vote data'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Loading data
Stata data files end in `*.dta`. They are easily loaded with the `use`
command. Some datasets, like the one we will use today, can be
downloaded from the Stata website directly with the `webuse`
command. While we won't really use these toy datasets in our research,
they can come in handy for small exercises (such as we will do
today). They are also the datasets used in the Stata help files.


```{r}
start <- 'load in school vote data'
end <- 'outsheet dataset'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## `outsheet` dataset
Stata can export data in almost any commonly-used format. The most commonly used
format for data files is in what's called ASCII delimited form, with a
delimiter of either a comma or a tab. To export a dataset we use the
`outsheet` command. You can specify which variables from the current dataset you'd like to
export. `*.csv` is a good format for Microsoft Excel files.

```{r}
start <- 'outsheet dataset'
end <- 'insheet dataset'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### QUICK EXERCISE
> Export the current dataset in tab delimited format, then go ahead and
> open up the new dataset in Excel.

<br>

## `insheet` dataset
Similarly, Stata can read in data in most any format using the
`insheet` command:

```{r}
start <- 'insheet dataset'
end <- 'describe data'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## `describe` dataset
Once you have your dataset in memory, you can `describe` it to get a
quick overview of what you have:

```{r}
start <- 'describe data'
end <- 'labeling data'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Labeling data and variables
Properly labeling datasets and variables will make life *much*
easier for you and anyone else who uses your dataset. To label an
entire dataset, use the `label data` command:

```{r}
start <- 'labeling data'
end <- 'labeling variables'
writeLines(logparse(lf, start = start, end = end))
```

To label variables, use the `label variable` command:


```{r}
start <- 'labeling variables'
end <- 'describe again'
writeLines(logparse(lf, start = start, end = end))
```
Make sure that your variable labels are highly descriptive and
directional (e.g., instead of labeling a binary variable **gender**
label it **female**). We can see our labels if once again `describe`
our data:

```{r}
start <- 'describe again'
end <- 'labeling values within variables'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Labeling values within variables
For many binary/categorical variables, you'll want to say what the
underlying levels of the variable in the dataset mean. This is a
two-part process. First you define the label values through `label define`; next you apply
them to your particular variable with `label values <variable>
<label>`:

```{r}
start <- 'labeling values within variables'
end <- 'transforming variables'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Transforming variables
Transforming a variable refers to using an operation to create a new
version of an existing variable. In our dataset, both income and
household spending on education are expressed in terms of the natural log
of the existing variable. If we want to have the values of these variables in nominal scale,
we need to use the `gen` command to create a new variable:

```{r}
start <- 'transforming variables'
end <- 'recoding variables'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## Recoding variables
Recoding a variable involves changing the values of a variable based
on its existing variables. We often want to recode variables in order
to make them more useful for whatever analysis we're working on. I'll
show you two different ways of recoding, one based on the `gen`
command and a more complicated one using the `egen` commend with the
`recode` command:

```{r}
start <- 'recoding variables'
end <- 'compute new variable'
writeLines(logparse(lf, start = start, end = end))
```

<br>

#### QUICK EXERCISE
> Create a variable that is equal to 1 if it's greater than the median of income. 
> Properly label the variable and its values.

<br>

## Computing a new variable
Computing a variable involves using the values of other variables to
create a new variable. For instance, to calculate an effective
property tax rate, we want to divide the property tax value by income:

```{r}
start <- 'compute new variable'
end <- 'end file'
writeLines(logparse(lf, start = start, end = end))
```

<br>

## EXERCISES
1. Create a new binary variable for whether or not the family has any children in public schools. Properly label your variable and variable values. 
2. Create a new variable for percent of household income spent on education. Properly label your new variable.
3. Create a new variable for persons with low, moderate and high percentages of spending on education. Label the variable and value labels properly.  
4. Tabulate household spending and voting for public school
funding. What do you find?

<br><br>

*Init: 06 June 2015; Updated: `r format(Sys.Date(), format = "%d %B %Y")`*

<br>
