\documentclass[12pt]{article}
\usepackage{bibentry}
\usepackage{fourier}
\usepackage[margin=1in]{geometry}

\usepackage[colorlinks=true,
                      pdfstartview=FitV,
                      urlcolor=blue,
]{hyperref}
\usepackage{natbib}



\begin{document}
\bibliographystyle{apalike}
\nobibliography{nature}

\thispagestyle{empty}%


\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}

\setcounter{secnumdepth}{-2}



\begin{flushleft}
Vanderbilt University\\
Leadership, Policy and Organizations\\
Class Number 9951\\ 
Fall 2018\\
\end{flushleft}

\begin{center}
\Large{\textbf{Methods Practicum}}\\
\end{center}

\begin{flushleft}
William R. Doyle\\
Office: 207D Payne\\
Office Hours: Tuesday and Thursday 2:30-4 or by Appointment \\
w.doyle@vanderbilt.edu\\
phone (615) 322-2904\\

\vskip 12 pt

\end{flushleft}

\section{Course Overview}%

The overview includes an introduction to the course, guidelines on grading, and required texts.

\subsection{Introduction}%
\begin{flushleft}

  This course is the first of a three semester series of courses
  designed to introduce you to the \textit{practice} of research,
  particularly the applied side of quantitative research. The goal of
  this course to help you to prepare a paper that can be presented at
  a major research conference and, hopefully, submitted to a journal
  for publication.

  To accomplish this goal, you will choose from among publicly
  available datasets. You will identify a research topic, then later a
  research question. You will create a dataset using the publicly
  available sources.  By the end of the semester, you will have a
  properly formatted and cleaned dataset, with auxiliary information
  from other sources included. Next semester we will analyze this
  dataset. By the end of May, you will complete a paper based on this
  analysis.

  Along the way, you will develop skills that will be helpful in
  future work using any kind of data. This class has a strong emphasis
  on using programming skills to aid in the replication of work and to
  simplify complex analyses.


\end{flushleft}

\subsection{Grading}%
\begin{flushleft}
Evaluation for the course will be based on the following factors:

\textit{Assignments: 50\%}

There will be weekly assignments,
pass/fail. Late assignments will not be accepted. These assignments
will account for half of your grade. Collaboration on assignments is
fine, however, many of the assignments will ask you to work with
variables and datasets of your own choosing.

\textit{Summary Paper and Codebook: 50\%}

At the end of the semester, you will need to present the results of
your data collection efforts, with a summary paper of no more than
five pages of text, accompanied by properly formatted tables and
graphics. The summary paper and codebook will be due on December
6. You will present your work to the rest of the class on December 6.



\end{flushleft}

\subsection{Texts}%

\begin{flushleft}

Baum, C (2006) \textit{An Introduction to Modern Econometrics Using
STATA}. College Station: STATA Press

Long, J.S. (2009) \textit{The Workflow of Data Analysis Using Stata}. College Station: STATA Press

For this semester, both of these books are \textit{optional}: I will
recommend a few chapters from both as the semester progresses, but the
class notes are the only required reading. 

\end{flushleft}

\subsection{Software}

You need to have access to a working version of Stata, (at least v
13.0).  Stata is installed on computers on Peabody campus, including
Wyatt 132, and on stations in the Peabody library. You are not
required to purchase Stata, but you will need to use it for class
assignments.

If you do purchase Stata, you will need Stata IC (standard
version). Vanderbilt has what's called a gradplan with Stata under
which you can purchase the software at greatly reduced prices.  Stata
SE is a more-powerful version of Stata that is useful for the larger
datasets many of you may be working with.


\subsection{Honor Code}
\label{sec:honor-code}

For this course, you are bound by the terms of the Peabody Honor
System. Any breach of academic honesty, including cheating,
plagiarism, or failing to report a known or suspected violation of the
Code will be reported to the Honor Council. In particular, papers must
assign credit to the sources you use. Material borrowed from
another--quotations, paraphrases, key words, or ideas--must be
credited following appropriate citation procedures (footnotes and
bibliography). As mentioned above, collaboration \textit{is} permitted
on assignments but \textit{is not} permitted on your summary paper and
codebook. 
 
If you have any doubts, please ask me for clarification. Uncertainty
about the application of the Honor Code does not excuse a violation.

\section{Schedule for Meetings}

The schedule for all class meetings is as follows:

\begin{flushleft}


\subsection{August 22}

Intro to Stata

\subsection{August 29}

\textit{Topics}:

Getting Started with Research Questions, Introduction to literate
programming in stata

\subsection{September 5}


\textit{Topics}:
STATA Basics

\textit {Assignments:}
Assignment 1 due September 2, midnight

\subsection{Septermber 12}

\textit{Topics}:

Working with publicly available datasets

\textit {Assignments:}
Assignment 2 due September 9, midnight

\subsection{September 19}


\textit{Topics}:

Dataset manipulation: Collapsing, merging, bending the data to your will


\textit {Assignments:}
Assignment 3 due September 16, midnight
\subsection{September 26}


\textit{Topics}:
More dataset manipulation: One to many merging, many to one merging, appending datasets

\textit {Assignments:}
Assignment 4 due September 23, midnight

Summary research area due

\subsection{October 3}


\textit{Topics}:

Sampling: Simple sampling designs

\textit {Assignments:}
Assignment 5 due October 2, midnight
\subsection{October 10}


\textit{Topics}:
Sampling: Complex sampling designs


\textit {Assignments:}
Assignment 6 due October 9, midnight
\subsection{October 17}


\textit{Topics}:

Data Cleaning

\textit {Assignments:}
Assignment 7 due October 16, midnight

\subsection{October 24}


\textit{Topics}:

Data Validation

\textit {Assignments:}
Assignment 8 due October 23, midnight

Research questions due. 

\subsection{October 31}


\textit{Topics}:

Descriptive Statistics: Tabular and Graphical Approaches

\textit {Assignments:}

Assignment 9 due October 28, midnight

\subsection{November 7}


\textit{Topics}:

An Introduction to Programming: Using Macros

\textit {Assignments:}

Assignment 10 due November 5, midnight

\subsection{November 14}

\textit{Topics}:

Further graphical approaches to describing data 

Assignment 11 due November 12, midnight

\subsection{November 21}

No Class, Thanksgiving Break


\subsection{November 28}


\textit{Topics}:

Topic TBA: based on student work. 


\subsection{December 5}

\textit{Topics}:

Class Presentations

\textit {Assignments:}

Summary Papers, Do-File and Codebooks due before class, December 5. 


\end{flushleft}
\end{document}




















